Arrows for Parallel Computations

### 2017-05-26 paper/content/motivation.tex
minor

HIT HERE REALLY STRONG

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours. and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \code{map} with and without load balancing, \code{map-reduce}) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\code{pipe}, \code{ring}, \code{torus}). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

>>>>

\olcomment{HIT HERE REALLY STRONG}

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
\end{itemize}

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel |map| with and without load balancing, |map-reduce|) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

### 2017-05-26 paper/content/parallelHaskells.tex
minor

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel, as also Figure~\ref{fig:parEvalN} symbolically shows:

\begin{code}
parEvalN :: [a -> b] -> [a] -> [b]
\end{code}

\begin{figure}[h]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of \texttt{parEvalN}.}
	\label{fig:parEvalN}
\end{figure}

Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with |using :: a -> Strategy a -> a|. For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form. To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations. This is particularly necessary in circular communication topologies like in the |torus| or |ring| skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without |pseq| during testing for this paper.

@@

\begin{figure}[h]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the Multicore Haskell parEvalN version}
\olcomment{Evaluation step explicitly shown?}
\end{figure} %$ %% formatting

@@

Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |DOLLAR| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|. This results in |Par [b]|. We finally execute this process with |runPar| to finally get |[b]| again.

>>>>

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of |parEvalN|.}
	\label{fig:parEvalN}
\end{figure}
Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|. For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form. To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations. This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.

@@

\begin{figure}[h]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{Dataflow of the Multicore Haskell parEvalN version}
\end{figure} %$ %% formatting

@@

Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|. This results in |Par [b]|. We finally execute this process with |runPar| to finally get |[b]| again.
